---
title: "Quantitative analysis of A-to-I RNA editing in mouse"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    theme: united
    highlight: tango
    df_print: paged
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE)
```

```{r}
BiocManager::install("TxDb.Mmusculus.UCSC.mm39.knownGene")
BiocManager::install("Mus.musculus")

library(here)
library(ggplot2)
library(tximport)
library(data.table)
library(ComplexHeatmap)
library(SummarizedExperiment)
library(viridis)
library(MetBrewer)
library(Mus.musculus)
library(TxDb.Mmusculus.UCSC.mm39.knownGene) 
library(rtracklayer)
library(raer)
library(raerdata)
library(Rsamtools)
library(BiocParallel)
library(GenomicFeatures)
library(ggbio)
library(cowplot)

TxDb(Mus.musculus) <- TxDb.Mmusculus.UCSC.mm39.knownGene
threads <- 24
theme_set(theme_cowplot())

gtf <- import.gff("/genome.gtf.gz")

if (!dir.exists("plots")) dir.create("plots")
```

```{r}
sample_order <- c("ADAR1p110-replica",
                  "ADAR1p150-replica",
                  "RFP_editase_neg_ctrl"
                  )

subset_to_plot <- c("ADAR1p110-replica",
                  "ADAR1p150-replica",
                  "RFP_editase_neg_ctrl"
                  )

subset_nice_labels <- c("ADAR1p110-replica",
                  "ADAR1p150-replica",
                  "RFP_editase_neg_ctrl"
                  )

##Number of Sample replicates# 
#For equal number of sample replicates ( e.g three samples per group)
#samples <- paste0(rep(sample_order, each = 3), "_", c("1", "2", "3"))

#For unequal number of sample replicates (e.g 3 p110_replica, 3 p150_replica, 2 neg_cntrl)
rep_counts <- c(3, 3, 1)
samples <- unlist(
  mapply(function(s, n) paste0(s, "_", seq_len(n)),
         sample_order,
         rep_counts,
         SIMPLIFY = FALSE)
)

sample_info <- data.frame(sample_id = factor(samples, levels = samples), 

  bam_fn = file.path("/data7/Kleinova/2024-04-04/star/",  

                     samples,  

                     paste0(samples, "_dedup_sorted.bam")), 

            salmon_fn = file.path("/data7/Kleinova/2024-04-04/salmon",  

                        samples,  

                        "quant.sf")) 

missing_salmon <- sample_info$salmon_fn[!file.exists(sample_info$salmon_fn)] 

missing_bam <- sample_info$bam_fn[!file.exists(sample_info$bam_fn)] 

  

missing_salmon 

missing_bam 


stopifnot(all(file.exists(sample_info$salmon_fn)))
```

## ADAR abundance levels

Shown below is a heatmap showing the estimated abundance of endogenous ADAR1 transcripts and the ADAR1 transgenes. 

```{r}
tx2gene <- as.data.frame(gtf)[ , c("gene_name", "transcript_id")]
tx2gene$transcript_id_no_version <- sub("\\.[0-9]+$", "", tx2gene$transcript_id)
tx2gene <- unique(tx2gene)

salmon_tx <- tximport(sample_info$salmon_fn, txOut = TRUE, type = "salmon")
rownames(salmon_tx$abundance) <- sub("\\.[0-9]+$", "", rownames(salmon_tx$abundance))

genes <- tx2gene$gene_name[match(rownames(salmon_tx$abundance),
                                 tx2gene$transcript_id_no_version)]

genes[is.na(genes)] <-  rownames(salmon_tx$abundance)[is.na(genes)]

adar_tx <- rownames(salmon_tx$abundance)[genes == "Adar"]

tpms <- salmon_tx$abundance

tpm_mat <- tpms[adar_tx, ]
colnames(tpm_mat) <- sub("/quant.sf", "", sample_info$salmon_fn) |>  basename()

sample_info$n_reads <- colSums(salmon_tx$counts)
```




```{r fig.height = 10, fig.width = 15}
set_color_palette <- function(vec){
  vec <- na.omit(vec)
  pal <- as.character(MetBrewer::met.brewer("Java", length(vec)))
  names(pal) <- vec
  pal
}

tableau20 <- c(
    "#1F77B4",
    "#AEC7E8",
    "#FF7F0E",
    "#FFBB78",
    "#2CA02C",
    "#98DF8A",
    "#D62728",
    "#FF9896",
    "#9467BD",
    "#C5B0D5",
    "#8C564B",
    "#C49C94",
    "#E377C2",
    "#F7B6D2",
    "#7F7F7F",
    "#C7C7C7",
    "#BCBD22",
    "#DBDB8D",
    "#17BECF",
    "#9EDAE5"
  )


Heatmap(log2(tpm_mat + 1)[, samples], 
        col = viridis(100),
        column_title = "ADAR1 transcript abundance", 
        name = "log2 TPM",
        column_names_gp = gpar(fontsize = 8),
        cluster_columns = FALSE,
        cluster_rows = FALSE)


y <- Heatmap(log2(tpm_mat + 1)[, samples], 
        col = viridis(100),
        column_title = "ADAR1 transcript abundance", 
        name = "log2 TPM",
        column_names_gp = gpar(fontsize = 8),
        cluster_columns = FALSE,
        cluster_rows = FALSE)

pdf("plots/chk5.pdf", y, width = 8, height = 8)
draw(y)
dev.off()
```

### Gene expression differences  

```{r, fig.width = 8, fig.height=6}
library(DESeq2)
tg_names <- tail(rownames(salmon_tx$abundance), 10)
t2g <- na.omit(tx2gene)
t2g <- data.frame(transcript_id = t2g$transcript_id,
                  gene_name = t2g$gene_name)

salmon_tx <- tximport(sample_info$salmon_fn, tx2gene = t2g, type = "salmon", ignoreTxVersion = TRUE)

pdata <- sample_info
pdata$condition <- sub("_[0-9]$", "", pdata$sample_id)
pdata <- pdata[, c("sample_id", "condition")]
pdata$sample_id <- make.names(pdata$sample_id)
pdata$condition <- make.names(pdata$condition)
rownames(pdata) <- pdata$sample_id
dds <- DESeqDataSetFromTximport(salmon_tx, pdata, ~condition)
dds <- DESeq(dds)
vst_tr <- vst(dds)

vst_mat <- assay(vst_tr)
n_genes <- 2000
var_genes <- order(rowVars(vst_mat), decreasing = T)[1:n_genes]
pc <- prcomp(t(vst_mat[var_genes, ]))

pc_res <- pc$x |> as.data.frame() 
pc_res <- cbind(id = rownames(pc_res), pc_res) 
pc_res$condition <- sub("_[0-9]$", "", pc_res$id)

pct_var <- scales::percent(pc$sdev^2 / sum(pc$sdev^2), accuracy = 0.01)

p <- ggplot(pc_res, aes(x = PC1, y = PC2, color = condition)) +
  geom_point() +
  scale_color_manual(values = tableau20) +
  labs(x = paste0("PC1 (", pct_var[1], ")"),
       y = paste0("PC2 (", pct_var[2], ")"),
       color = NULL)
p
```

A plot expression of a few Interferon stimulated genes (and MDA5).

```{r, fig.width = 8}
dds$condition_ordered <- factor(dds$condition, levels = make.names(sample_order))

plot_counts <- function(dds, gene) {
  p <- plotCounts(dds, gene, intgroup = "condition_ordered", returnData = TRUE) |>
    ggplot(aes(condition_ordered, count)) + 
    geom_jitter() +
    labs(x = NULL,
         y = "Normalized counts",
         subtitle = gene) +
    theme(axis.text.x = element_text(angle = 90,
                                     hjust = 1, 
                                     vjust = 0.5))
    p
}
```

```{r, fig.width = 6, fig.height = 60}
#Mouse specific gene names
plts <- lapply(c("Ifnb1", "Isg15", "Ifit2", "Ifih1", "Ddx58", "Ddx39b", "Oas2", "Ifna2", "Ifna11", "Ifnar1", "Ifnar2", "Eif2ak2", "Zbp1", "Adar", "Adarb1", "Adarb2"),
 plot_counts, dds = dds)

genes_plot <- plot_grid(plotlist = plts, ncol = 1, nrow = 16)
pdf("plots/chk9.pdf", width = 6, height = 50)
print(genes_plot)  
dev.off()

plot_grid(plotlist = plts, ncol = 1, nrow = 16)
```

```{r,  eval = FALSE}
# get IFN related genes

library(ReactomePA)
library(org.Mm.eg.db)

# get Reactome gene set for interferon signaling (mouse: R-MMU-913531)
gs <- gson_Reactome("mouse")
eids <- gs@gsid2gene[gs@gsid2gene$gsid == "R-MMU-913531", ]$gene

# convert to mouse SYMBOLs
ifn_genes <- select(org.Mm.eg.db, 
                    eids, 
                    columns = "SYMBOL",
                    keytype = "ENTREZID")$SYMBOL

ifn_mat <- vst_mat[intersect(rownames(vst_mat), ifn_genes), ]
ifn_mat <- t(scale(t(ifn_mat)))
ifn_mat[ifn_mat > 3] <- 3
ifn_mat[ifn_mat < -3] <- -3
#Replace NA/NaN with 0
ifn_mat[!is.finite(ifn_mat)] <- 0
colnames(ifn_mat) <- colnames(dds)

# doesn't really help with interpretation
Heatmap(ifn_mat,
        show_column_dend = FALSE,
        show_row_dend = FALSE,
        row_names_gp = gpar(fontsize = 5),
        name = "Zscore (deseq2-zst)",
        col = viridis(100),
        column_title = "IFN genes", 
        column_names_gp = gpar(fontsize = 8),
        column_labels = dds$condition_ordered,
        height = unit(10, "in"))

#save pdf
p <- Heatmap(ifn_mat,
        show_column_dend = FALSE,
        show_row_dend = FALSE,
        row_names_gp = gpar(fontsize = 5),
        name = "Zscore (deseq2-zst)",
        col = viridis(100),
        column_title = "IFN genes", 
        column_names_gp = gpar(fontsize = 8),
        column_labels = dds$condition_ordered,
        height = unit(10, "1n"))
pdf("plots/chk10.pdf", width = 8, height = 10)
print(p)
dev.off()
```
## RNA editing

Examine A-to-I editing sites. Sites from the rediportal database will be quantified.

```{r}
fafn <- "/genome.fa"

outfn_se <- "REDIportal_rse.rds"

if(!file.exists(outfn_se)){
  stopifnot(all(file.exists(sample_info$bam_fn)))
  ## Load REDIportal file
  rp <- readRDS("/rediportalll__mm39_rse.rds")
  bf <- BamFile(sample_info$bam_fn[1])
  chroms <- seqnames(keepStandardChromosomes(seqinfo(bf)))
  chroms <- intersect(unique(seqnames(rp)), chroms)
  rp <- rp[seqnames(rp) %in% chroms]
  rp <- keepSeqlevels(rp, chroms)
  bpp <- MulticoreParam(workers = threads, progressbar = TRUE)

  
  fp <- FilterParam(max_depth = 1e6,
                  min_base_quality = 30,
                  min_mapq = 255,
                  library_type = "fr-first-strand",
                  trim_5p = 5L,
                  trim_3p = 5L)
  
  bam_fns <- sample_info$bam_f
  names(bam_fns) <- sample_info$sample_id
  
  rse <- pileup_sites(bam_fns,
                      fasta = fafn,
                      sites = rp, 
                      chroms = chroms,
                      param = fp,  
                      BPPARAM = bpp,
                      verbose = TRUE)
   
   saveRDS(rse, outfn_se)
   bpstop(bpp)
} else {
  rse <- readRDS(outfn_se)
}

```

```{r}

if(!file.exists("aei_vals_mouse.rds")) {
  ## 1. Load RepeatMasker annotations for mouse
  if(!file.exists("rmsk_mm39.rds")){
    library(AnnotationHub)
    ah <- AnnotationHub()
    query(ah, c("RepeatMasker", "Mus musculus", "mm39"))
    rmsk_mm39 <- ah[["AH99013"]]   # <- pick the correct AH id for mm39 RepeatMasker #AH99013
    saveRDS(rmsk_mm39, "rmsk_mm39.rds")
  } else {
    rmsk_mm39 <- readRDS("rmsk_mm39.rds")
  }
  
  ## 2. Select SINEs in mouse (B1, B2, B4, ID)
  mouse_sines <- rmsk_mm39[rmsk_mm39$repClass %in% c("SINE")]
  mouse_sines <- keepStandardChromosomes(mouse_sines, pruning.mode = "coarse")
  mouse_sines <- mouse_sines[!seqnames(mouse_sines) %in% c("chrM", "chrY")]
  
  ## 3. Overlap with genic regions
  library(TxDb.Mmusculus.UCSC.mm39.knownGene)
  gene_gr <- genes(TxDb.Mmusculus.UCSC.mm39.knownGene)
  genic_sines <- subsetByOverlaps(mouse_sines, gene_gr, ignore.strand = TRUE)
  genic_sines <- GenomicRanges::reduce(genic_sines)
 
  seqlevelsStyle(genic_sines) <- "NCBI"
  genome(genic_sines) <- NA

  ## 4. Restrict to chroms present in BAM
  chroms_in_bam <- seqnames(seqinfo(BamFile(sample_info$bam_fn[1])))
  chroms_in_both <- intersect(unique(seqnames(genic_sines)), chroms_in_bam)
  genic_sines <- genic_sines[seqnames(genic_sines) %in% chroms_in_both, ]
  
  
  ## 5. Parallel setup
  bpp <- MulticoreParam(workers = threads, progressbar = TRUE)

  
  ## 6. SNP annotation for mouse (dbSNP mm39, custom BED)
library(rtracklayer)
library(GenomicRanges)

if(!file.exists("mm39_snps_genic_sines_fol.bed.gz")) {
snp_bed_file <- "/mouse_snps_merged.sorted.bed"

# Import SNPs from the BED file
snps_gpos <- import(snp_bed_file, format = "BED")
seqlevelsStyle(genic_sines) <- "NCBI"
genome(genic_sines) <- NA
seqlevels(snps_gpos)[seqlevels(snps_gpos) == "chrMT"] <- "MT"
seqlevelsStyle(snps_gpos) <- "NCBI"


###Alternative!!
hits <- findOverlaps(genic_sines, snps_gpos)
snps_fol <- snps_gpos[unique(subjectHits(hits))]
#refine structure to gpos
snps_fol_gpos <- resize(snps_fol, width = 1)
mcols(snps_fol_gpos)$name <- "."
mcols(snps_fol_gpos)$score <- 0
strand(snps_fol_gpos) <- "*"


export(snps_fol_gpos, "mm39_snps_genic_sines_fol.bed.gz")
} else {
snps_fol_gpos <- import.bed("mm39_snps_genic_sines_fol.bed.gz")
 }


   ## 8. Run AEI calculation
  bam_fns <- sample_info$bam_fn
  names(bam_fns) <- sample_info$sample_id

  ###### continue AEI calculation##########################

  aei_vals <- calc_AEI(bam_fns,
                       fasta = fafn,
                       alu_ranges = genic_sines,   # <-- replaced with mouse SINEs
                       snp_db = snps_fol_gpos,
                       param = FilterParam(min_mapq = 255,
                                           trim_5p = 5,
                                           trim_3p = 5,
                                           min_base_quality = 30),
                       BPPARAM = bpp,
                       verbose = TRUE)

  saveRDS(aei_vals, "aei_vals_mouse.rds")
} else {
  aei_vals <- readRDS("aei_vals_mouse.rds")
}

# Peek at the first element
head(aei_vals[[1]])  

```

```{r}
#extract A-G editing values per chromosome
library(dplyr)
library(tidyr)
library(writexl)

df <- aei_vals$AEI_per_chrom
df_ag <- df %>%
  filter(allele == "A_G")
df_wide <- df_ag %>%
  select(bam_file, chrom, prop) %>%
  pivot_wider(names_from = chrom, values_from = prop)
write_xlsx(df_wide, "AEI_AtoG_per_chromosome_per_sample.xlsx")
```


```{r, fig.width = 8, fig.height = 12, fig.cap = "Shown are the number of editing sites from RADAR that are detected (>= 1 read supporting site) in each sample. The data shown in the left panel are normalized to account for the number of sequencing reads in each sample, whereas the panel on the right shows the unnormalized number of sites."}

library(cowplot)
v <- assay(rse, "ALT")
r <- rowData(rse)$REF
v <- as.data.table(v)
v$site_id <- rownames(rse)
v$REF <- r
vdf <- data.table::melt.data.table(as.data.table(v), 
                                   measure.vars = 1:(ncol(v) - 2))
vdf <- vdf[value != "-"]
vdf$var_allele <- paste0(vdf$REF, vdf$value)
vdf <- vdf[vdf$var_allele == "AG"]
vdf <- vdf[, .(n_sites = .N), by = variable]

vdf <- merge(vdf, setDT(sample_info)[,.(sample_id, n_reads)], by.x = "variable", by.y = "sample_id")
vdf$norm_n_sites <- 10e6 * vdf$n_sites / vdf$n_reads
vdf$sites_per_million <- 1e6 * vdf$n_sites / vdf$n_reads


```

```{r, fig.height= 16, fig.width= 20}
pa <- ggplot(vdf, aes(forcats::fct_rev(variable), sites_per_million)) + 
geom_col(fill = "navy") +  coord_flip() +
  labs(
    x = NULL,
    y = "# of editing sites\n(per million reads)")


pl <- ggplot(vdf, aes(forcats::fct_rev(variable), norm_n_sites)) + 
  geom_col(fill = "navy") +
  coord_flip() +
  labs(
    x = NULL,
    y = "Total # of editing sites") +
    theme(
    #axis.title = element_text(size = 14),
    axis.text = element_text(size = 16)
  )


grid_plot <- plot_grid(pa, pl, 
	  #labels = c("A", "B"), 
          align="v",
          rel_widths = c(6),
          rel_heights = c(30),
          nrow = 5)

ggsave("plots/chkkkk.pdf", plot = grid_plot, width = 20, height = 30, units = "in")

pa
pl

```


```{r}
vdf$condition <-  factor(sub("_[0-9]$", "", as.character(vdf$variable)), levels = sample_order)
vdf

p <- ggplot(vdf, aes(forcats::fct_rev(condition), sites_per_million)) + 
  geom_jitter(width = 0.2) +
    stat_summary(
    fun.y = "mean",
    col = "red",
    size = 0.2,
    shape = 3
  ) + 
  coord_flip() +
  labs(
    x = NULL,
    y = "# of editing sites\n(per million reads)")

pp <- ggplot(vdf[condition %in% subset_to_plot], aes(condition, sites_per_million)) + 
  geom_jitter(width = 0.2) +
    stat_summary(
    fun.y = "mean",
    col = "red",
    size = 0.2,
    shape = 3
  ) + 
  labs(
    x = NULL,
    y = "# of editing sites\n(per million reads)") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))


pr <- ggplot(vdf[condition %in% subset_to_plot], aes(condition, norm_n_sites)) + 
  geom_jitter(width = 0.2) +
    stat_summary(
    fun.y = "mean",
    col = "red",
    size = 0.2,
    shape = 3
  ) + 
  labs(
    x = NULL,
    y = "Total # of editing sites") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))

p
pp
pr

save_plot("plots/chk15_1.pdf", p, base_asp = 1, base_height = 6)
save_plot("plots/chk15_2.pdf", pp, base_asp = 1, base_height = 4)
save_plot("plots/chk15_3.pdf", pr, base_asp = 1, base_height = 4)

```

```{r, eval = FALSE, fig.cap="Editing frequency (# of edited reads / total reads) of REDIPORTAL sites across each sample."}
rse02 <- calc_edit_frequency(rse,
    edit_from = "A",
    edit_to = "G",
    drop = TRUE
)

rse02
```

```{r, fig.cap="Editing frequency (# of edited reads / total reads) of REDIPORTAL sites across each sample."}
rse <- calc_edit_frequency(rse)
plt_dat <- assay(rse, "edit_freq") |>
  as.data.table() 
plt_dat <- cbind(id = rownames(plt_dat), plt_dat ) |>
  melt(id.vars = "id")
plt_dat$variable <- factor(plt_dat$variable, levels = levels(sample_info$sample_id))
plt_dat$condition <-  factor(sub("_[0-9]$", "", as.character(plt_dat$variable)), levels = sample_order)

N <- ggplot(plt_dat, aes(variable, value)) + 
  geom_boxplot(width = 0.2,
               fill="white",
               outlier.shape = NA)  +
  labs(x = NULL,
       y = "Editing frequency") +
  theme(axis.text.x = element_text(angle = 90, 
                                   hjust = 1, 
                                   vjust = 0.5))

save_plot("plots/Editing frequency.pdf", N, base_asp = 1, base_height = 4)

N

```


```{r, fig.cap="Editing frequency (# of edited reads / total reads) of all REDIPORTAL sites summed across each sample."}
alt <- assay(rse, "nG")
ref <- assay(rse, "nA")
edit_idx <- colSums(alt) / (colSums(alt) + colSums(ref))
edit_idx <- data.frame(sample_id = factor(names(edit_idx), 
                                          levels = levels(sample_info$sample_id)),
                       edit_idx = edit_idx)
edit_idx$condition <-  factor(sub("_[0-9]$", "", as.character(edit_idx$sample_id)), levels = sample_order)

edit_idx
```

```{r}
pa <- ggplot(edit_idx, aes(condition, edit_idx)) + 
  geom_jitter(width = 0.2) + 
  stat_summary(
    fun.y = "mean",
    col = "red",
    size = 0.2,
    shape = 3
  ) + 
  labs(
    x = NULL,
    y = "Editing index (based on Rediportal sites)") +
  theme(axis.text.x =element_text(angle = 90, hjust = 1, vjust = 0.5))
pa
save_plot("plots/chk19.pdf", pa, base_asp = 1, base_width = 4)

```


```{r, fig.height = 25, fig.width = 10}
#ggplot of log-transformed editing index (edit_idx)
library(ggplot2)
library(ggpubr)

A <- ggplot(edit_idx, aes(x = condition, y = edit_idx)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.5) +
  theme_bw() +
  labs(y = "Editing Index", x = "Condition")

B <- ggplot(edit_idx, aes(x = condition, y = abs(log2(edit_idx)))) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.5) +
  theme_bw() +
  labs(y = "Abs.Log2 Editing Index", x = "Condition")

C <- ggplot(edit_idx, aes(x = condition, y = log2(edit_idx))) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.5) +
  theme_bw()+
  labs(y = "Log2 Editing Index", x = "Condition")


grid_plot <- plot_grid(A, B, C,  
          labels = c('A', 'B', 'C'),
          align="v",
          rel_widths = c(6),
          rel_heights = c(20),
          nrow = 6)
ggsave("plots/chk20.pdf", plot = grid_plot, width = 12, height = 20, units = "in")


plot_grid(A, B, C,   
          labels = c('A', 'B', 'C'),
          align="v",
          rel_widths = c(6),
          rel_heights = c(30),
          nrow = 6)
```


```{r}
p <- ggplot(edit_idx[edit_idx$condition %in% subset_to_plot,  ], aes(condition, edit_idx)) + 
  geom_jitter(width = 0.1) + 
  stat_summary(
    fun.y = "mean",
    col = "red",
    size = 0.2,
    shape = 3
  ) + 
  labs(
    x = NULL,
    y = "Editing index (REDIportal)") +
  theme(axis.text.x =element_text(angle = 90, hjust = 1, vjust = 0.5))

save_plot("plots/chk21.pdf", p, base_asp = 1, base_height = 4)

p
```

```{r, fig.height = 8, fig.width = 8, fig.cap="The ALU editing index (AEI) is a measure of the amount of editing occurring at ALU elements. It tends to be a more robust measure of editing than using the editing frequency of known sites. Higher values indicate more editing, and can be interpreted as a weighted average of the percent of edited bases in ALU repeats"}
library(ggrepel)
aei_vals_plt <- as.data.table(aei_vals$AEI)
aei_vals_plt$sample_id <- factor(rownames(aei_vals$AEI), levels = levels(sample_info$sample_id))
aei_vals_plt <- melt(aei_vals_plt, id.vars = "sample_id")

aei_vals_plt$name <- sub( "_", "_to_", aei_vals_plt$variable)

aei_vals_plt <-  aei_vals_plt[aei_vals_plt$sample_id %in% sample_info$sample_id, ]

aei_vals_plt <- aei_vals_plt[aei_vals_plt$variable == "A_G"]

aei_vals_plt$condition <- sub("_[0-9]$", "", as.character(aei_vals_plt$sample_id))
aei_vals_plt$condition <- factor(aei_vals_plt$condition, levels = sample_order)


p <- ggplot(aei_vals_plt, aes(forcats::fct_rev(condition), value)) +
  geom_jitter() +
  coord_flip() + 
  labs(x = NULL,
       y = "Editing index (AEI)") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
save_plot("plots/chk22_1.pdf", p, base_asp = 1, base_width = 4)
p


pv <- ggplot(aei_vals_plt, aes(forcats::fct_rev(condition), value)) +
  geom_jitter() +
  stat_summary(
    fun.y = "mean",
    col = "red",
    size = 0.2,
    shape = 3
  ) + 
  coord_flip() + 
  labs(x = NULL,
       y = "Editing index (AEI)") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
save_plot("plots/chk22_2.pdf", pv, base_asp = 1, base_width = 4)
pv


C <- ggplot(aei_vals_plt, aes(forcats::fct_rev(condition), x = condition, y = log2(value))) +
  geom_jitter() +
  coord_flip() + 
  labs(x = NULL,
       y = "log2(Editing Frequency + 1)") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))

save_plot("plots/chk22_3.pdf", C, base_asp = 1, base_height = 4)
C

aei_vals_plt
```

```{r}
p <- ggplot(aei_vals_plt[aei_vals_plt$condition %in% subset_to_plot], aes(condition, value)) +
  geom_jitter(width = 0.1) +
    stat_summary(
    fun.y = "mean",
    col = "red",
    size = 0.2,
    shape = 3
  ) + 
  labs(x = NULL,
       y = "Editing index (AEI)") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))

p
save_plot("plots/chk23_1.pdf", p, base_asp = 1, base_height = 4)

p <- ggplot(aei_vals_plt[aei_vals_plt$condition %in% subset_to_plot], aes(condition, log2(value))) +
  geom_jitter(width = 0.1) +
    stat_summary(
    fun.y = "mean",
    col = "red",
    size = 0.2,
    shape = 3
  ) + 
  labs(x = NULL,
       y = "log2(Editing index (AEI))") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))

p
save_plot("plots/chk23_2.pdf", p, base_asp = 1, base_height = 4)
```

```{r, fig.height = 8, fig.width = 12, fig.cap="The AEI, as shown above, but also plotting other types of mismatches (which shouldn't have signal)"}

# Peek at the first element
head(aei_vals[[1]])    

# Summary across all samples
sum(sapply(aei_vals, nrow))   # total sites detected


y <- Heatmap(aei_vals$AEI, 
         name = "Editing index (AEI)",
         cluster_rows = FALSE,
         row_names_side = "left",
         col = viridis(100),
         row_names_gp = gpar(fontsize = 8),
        width = unit(6, "inch"))

pdf("plots/chk24.pdf", width = 8, height = 10)
print(y)
dev.off()

```


```{r, fig.height=12, fig.width = 20, fig.cap = "editing per Chromosome per sample"}
library(dplyr)
library(tidyr)
library(ComplexHeatmap)
library(circlize)
library(viridis)
library(grid)

aei_vals <- readRDS("aei_vals_mouse.rds")
df <- aei_vals$AEI_per_chrom

df_ag <- df %>%
  filter(allele == "A_G")
heatmap_df <- df_ag %>%
  select(bam_file, chrom, prop) %>%
  pivot_wider(names_from = chrom, values_from = prop)

# Convert to matrix
mat <- as.data.frame(heatmap_df)
rownames(mat) <- mat$bam_file
mat$bam_file <- NULL
mat <- as.matrix(mat)
mat_scaled <- t(scale(t(mat)))


y <- Heatmap(mat_scaled,
             name = "A-G editing per chrom",
             cluster_rows = FALSE,
             row_names_side = "left",
             col = viridis(100),
             row_names_gp = gpar(fontsize = 8),
             width = unit(8, "inch"))

pdf("plots/APChrom_scaled.pdf", width = 10, height = 10)
print(y)
dev.off()


Heatmap(mat_scaled,
             name = "A-G editing per chrom",
             cluster_rows = FALSE,
             row_names_side = "left",
             col = viridis(100),
             row_names_gp = gpar(fontsize = 8),
             width = unit(8, "inch"))
```


## Hyperediting Detection

```{r}
data_dir <- ("/2024-04-04/hyperediting/bed_filtered/")
col_vals <- MetBrewer::met.brewer("Java", 12) 

alleles  <- list(
              "A_to_C",
              "A_to_G",
              "A_to_T",
              "C_to_A",
              "C_to_G",
              "C_to_T",
              "G_to_A",
              "G_to_C",
              "G_to_T",
              "T_to_A",
              "T_to_C",
              "T_to_G")

read_hyperedits <- function(allele, data_dir, type = "reads"){
  # first load in bed files of individual hyperedited reads
  files <- dir(file.path(data_dir, type, allele),
           pattern = "*.bed",
           recursive = T,
           full.names = T)

  dat <- lapply(files, function(x) {
    if(file.info(x)$size == 0) {
      return(NULL)
    }
    fread(x, col.names = c("chrom",
                           "start",
                           "end",
                           "id",
                           "mismatches",
                           "strand"))
  })

  names(dat) <- basename(files)

  dat <- rbindlist(dat, idcol = "name")
  # exit if no sites/reads exist
  if (nrow(dat) == 0){ return(NULL) }

  dat$sample <- sub("_[12]_[AGCT]to[AGCT]_gnome[ATCG]to[ATCG].bed$",
                    "", 
                    dat$name)
  dat
}
```

```{r}
edit_dat <- lapply(alleles, read_hyperedits, data_dir, type = "edits")

names(edit_dat) <- alleles
edit_dat <- rbindlist(edit_dat, idcol ="allele")
 
edit_dat$condition <- sub("_[0-9]$", "", as.character(edit_dat$sample))
#edit_dat[, condition := sub("_[0-9]$", "", as.character(sample))]
fwrite(edit_dat, "hyperediting_sites.tsv.gz")
```

```{r}
edit_dat[, allele := factor(allele, levels = alleles)]
edit_dat[, sample := factor(sample, levels = levels(sample_info$sample_id))]
edit_dat_summary <- edit_dat[, .(supporting_reads = length(unique(id))), 
                             by = .(allele,
                                    strand,
                                    chrom,
                                    start,
                                    end,
                                    sample)][,
                                             site_id := paste(chrom, 
                                                              start, 
                                                              strand, 
                                                              sep = ":")
                                             ]

edit_dat_summary
```

A-to-G hyperediting sites will be re-quantified with original BAM to generate counts of A and G in normally mapped reads overlapping the hyperedited sites. 

```{r}
he_gr <- edit_dat[edit_dat$allele == "A_to_G", 
                  .(chrom, start, end, strand)] |>
  makeGRangesFromDataFrame(starts.in.df.are.0based = TRUE) |>
  sort() |>
  unique()

outfn_se <- "he_rse.rds"

if(!file.exists(outfn_se)){
  stopifnot(all(file.exists(sample_info$bam_fn)))
  
  bf <- BamFile(sample_info$bam_fn[1])
  chroms <- seqnames(keepStandardChromosomes(seqinfo(bf)))
  chroms <- intersect(unique(seqnames(he_gr)), chroms)
  he_gr <- he_gr[seqnames(he_gr) %in% chroms]
  he_gr <- keepSeqlevels(he_gr, chroms)
  bpp <- MulticoreParam(workers = threads, progressbar = TRUE)

  fp <- FilterParam(max_depth = 1e6,
                  min_base_quality = 30,
                  min_mapq = 255,
                  library_type = "fr-first-strand",
                  trim_5p = 5L,
                  trim_3p = 5L)
  
  bam_fns <- sample_info$bam_f
  names(bam_fns) <- sample_info$sample_id
  
  he_rse <- pileup_sites(bam_fns,
                      fasta = fafn,
                      sites = he_gr, 
                      chroms = chroms,
                      param = fp,  
                      BPPARAM = bpp,
                      verbose = TRUE)
   
   saveRDS(he_rse, "he_rse.rds")
   bpstop(bpp)
} else {
  he_rse <- readRDS(outfn_se)
}
```


```{r, fig.height=9, fig.width= 9, fig.cap = "number of hyperediting sites"}
n_samples <- length(unique(edit_dat_summary$sample))

edit_dat_summary$condition <- sub("_[0-9]$", "", as.character(edit_dat_summary$sample))
edit_dat_summary$condition <- factor(edit_dat_summary$condition, levels = sample_order)

#col_pal <- c(scales::brewer_pal(palette = "Paired")(12), 
#            scales::viridis_pal(option = "H")(n_samples - 12))


if (n_samples <= 10) {
  col_pal <- scales::brewer_pal(palette = "Paired")(n_samples)
} else {
  col_pal <- c(
    scales::brewer_pal(palette = "Paired")(10),
    scales::viridis_pal(option = "H")(n_samples - 10)
  )
}

edit_dat_summary <- edit_dat_summary[, .N, by = .(sample, condition, allele)]  

plt_summary <- merge(edit_dat_summary, setDT(sample_info)[,.(sample_id, n_reads)], by.x = "sample", by.y = "sample_id")
plt_summary$norm_he_sites <- 10e6 * plt_summary$N / plt_summary$n_reads
length(unique(plt_summary$sample))
p <- plt_summary |>
  ggplot(aes(allele, norm_he_sites)) +
  geom_jitter(aes(color = condition)) +
  coord_flip() + 
  labs(y = "# of hyper-edited sites\n(per million mapped reads)",
       x = "variant type",
       fill = NULL) +
   scale_color_manual(values = col_pal)

p 
save_plot("plots/he_site_variant.pdf", p)
```


```{r}
p <- plt_summary[allele == "A_to_G", ] |>
  
  ggplot(aes(forcats::fct_rev(condition), norm_he_sites)) +
  geom_jitter(width = 0.1) +
  coord_flip() + 
  stat_summary(
    fun.y = "mean",
    col = "red",
    size = 0.2,
    shape = 3
  ) + 
  labs(y = "# of hyper-edited sites\n(per million reads)",
        x = NULL) +
   scale_color_manual(values = col_pal) +
  theme(legend.position = "none")

p 

save_plot("plots/chk30.pdf", p, base_asp = 1, base_height = 4)
```

```{r}
p <- plt_summary[allele == "A_to_G" & plt_summary$condition %in% subset_to_plot, ] |>
  ggplot(aes(condition, norm_he_sites)) +
  geom_jitter(width = 0.1) +
  stat_summary(
    fun.y = "mean",
    col = "red",
    size = 0.2,
    shape = 3
  ) + 
  labs(y = "# of hyper-edited sites\n(per million reads)",
       x = NULL) +
   scale_color_manual(values = col_pal) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
p
save_plot("plots/chk31.pdf", p, base_asp = 1, base_height = 4)
```

```{r, fig.height = 8, fig.width = 12, fig.cap="Number of hyperediting sites per sample as a heatmap"}
hmat <-  dcast(edit_dat_summary, ... ~ allele, fill = 0) |>
  as.data.frame()

#Remove NA sample rows
hmat <- hmat[!is.na(hmat$sample), ]

rownames(hmat) <- hmat$sample
hmat$sample <- NULL
hmat$condition <- NULL
 h <- Heatmap(hmat, 
         name = "Hyperediting sites",
         cluster_rows = FALSE,
         row_names_side = "left",
         col = viridis(100),
         row_names_gp = gpar(fontsize = 8),
        width = unit(6, "inch"))

pdf("plots/heHeatmap1.pdf", h, width = 8, height = 10)
print(h)
dev.off()

Heatmap(hmat, 
         name = "Hyperediting sites",
         cluster_rows = FALSE,
         row_names_side = "left",
         col = viridis(100),
         row_names_gp = gpar(fontsize = 8),
        width = unit(6, "inch"))


```


```{r, fig.cap = "correlation (Pearson) between AEI values and # of hyperediting sites. The right panel has log10 transformed the # of hyperediting sites."}
aeis <- aei_vals$AEI[, "A_G"]
he_vals <- hmat[, "A_to_G"]
names(he_vals) <- rownames(hmat)

plt_dat <- data.frame(aei_values = aeis,
           he_sites = he_vals[names(aeis)]) 

cor_linear <- signif(cor(plt_dat$aei_values, plt_dat$he_sites), 2)
p1 <- ggplot(plt_dat, aes(aei_values, he_sites)) + 
  geom_point() +
    labs(x = "AEI",
         y = "hyperediting sites") +
    annotate(x = 0.2, y = 0.9 * max(plt_dat$he_sites),
             geom = "text", 
             label = paste0("r = ", cor_linear))


cor_log <- signif(cor(plt_dat$aei_values, log10(plt_dat$he_sites)), 2)
p2 <- ggplot(plt_dat, aes(aei_values, he_sites)) + 
  geom_point() +
    scale_y_log10() + 
    labs(x = "AEI",
         y = "hyperediting sites") + 
    annotate(x = 0.2, y = 0.9 * max(plt_dat$he_sites),
             geom = "text", 
             label = paste0("r = ", cor_log))

plot_grid(p1, p2, ncol = 2, nrow = 1)



 
grid_plot <- plot_grid(p1, p2,  
          labels = c('A', 'B'),
          ncol = 2, nrow = 1)

#ggsave("plots/chk33.pdf", plot = grid_plot)

ggsave("plots/chk33.pdf", plot = grid_plot, width = 8, height = 8, units = "in")


```

```{r}
edit_dat$condition <- sub("_[0-9]$", "", edit_dat$sample)

edit_dat[, c("cl_start", "cl_end") := tstrsplit(mismatches, ":")]

he_sites <- edit_dat[, .(n_mismatches = length(unique(start))), by = .(chrom, cl_start, cl_end, strand, condition)]

p150_conds  <- c("ADAR1p110-replica", "ADAR1p150-replica")
he_sites <- he_sites[condition %in% p150_conds]
he_sites <- makeGRangesFromDataFrame(he_sites, 
                                     keep.extra.columns = TRUE,
                                     starts.in.df.are.0based = TRUE)

merged_clusters <- reduce(he_sites, with.revmap = TRUE)
# aggregate(he_sites, merged_clusters$revmap, n_mismatches = sum(n_mismatches))

p110_he_sites <- he_sites[he_sites$condition == "ADAR1p110-replica" ]
p150_he_sites <- he_sites[he_sites$condition == "ADAR1p150-replica" ]

p150_specific <- setdiff(p150_he_sites, p110_he_sites)
rtracklayer::export.bed(p150_specific, "p150_he_clusters.bed")
```

## Differential editing analysis

Allelic counts (e.g. A and Gs) for hyperediting sites were computed separately. Many of the hyperediting sites identified are also in REDIportal. To accurately examine the distribution of editing sites across repeats, the hyperediting counts will be merged in with the "G" counts from the REDIportal database. To handle novel sites not in REDIportal, the allelic (A and G) counts for these sites have been recomputed to ensure that we have accurate allelic counts of both traditionally aligned and hyperedited reads overlapping the novel sites. 


```{r}
if(!file.exists("rse-final.rds")) {
  he <- edit_dat[allele == "A_to_G"] |>
    makeGRangesFromDataFrame(starts.in.df.are.0based = TRUE, 
                             keep.extra.columns = TRUE)
  he$site_ids <-  paste(
    "site",
    decode(seqnames(he)),
    decode(start(he)),
    as.integer(strand(he)),
    sep = "_"
  )
  
  # make a matrix of read counts for HE sites per sample
  he_mat <- mcols(he) |> 
    as.data.table() |>
    dcast(site_ids ~ sample,
          value.var = "site_ids", 
          fun.aggregate = function(x) length(unique(x))) |>
    as.data.frame()
  
  rownames(he_mat) <- he_mat$site_ids
  he_mat$site_ids <- NULL
  he_mat <- as.matrix(he_mat)
  
  non_he_sites <- setdiff(rownames(rse), rownames(he_mat))
  rp_he_sites <- intersect(rownames(rse), rownames(he_mat))
  novel_he_sites <- setdiff(rownames(he_mat), rownames(rse))
  stopifnot(all(colnames(rse) %in% colnames(he_mat)))
  
  # extract novel HE sites from he_rse and add in HE counts
  stopifnot(all(colnames(he_rse) %in% colnames(he_mat)))
  
  novel_he_sites <- intersect(rownames(he_rse), novel_he_sites)
  
  he_rse <- he_rse[novel_he_sites, colnames(rse)]
  novel_he_mat <- he_mat[novel_he_sites, colnames(rse)]
  
  stopifnot(all(rownames(assay(he_rse, "nAlt")) == rownames(novel_he_mat)))
  stopifnot(all(colnames(he_rse) == colnames(novel_he_mat)))
  
  assay(he_rse, "nAlt") <- assay(he_rse, "nAlt") + novel_he_mat
  assay(he_rse, "nG") <- assay(he_rse, "nG") + novel_he_mat
  
  # extract HE sites that are also in rediportal
  rp_he_mat <- he_mat[rp_he_sites, colnames(rse)]
  rp_he_rse <- rse[rp_he_sites, ]
  
  stopifnot(all(rownames(assay(rp_he_rse, "nAlt")) == rownames(rp_he_mat)))
  stopifnot(all(colnames(rp_he_rse) == colnames(rp_he_mat)))
  
  assay(rp_he_rse, "nAlt") <- assay(rp_he_rse, "nAlt") + rp_he_mat
  assay(rp_he_rse, "nG") <- assay(rp_he_rse, "nG") + rp_he_mat
  
  nh_rse <- rse[non_he_sites, ]
  
  # I can't use rbind on these summarizedExperiemtns for some reason.
  # get a names don't match error ?
  # I'll just do it manually
  
  #calculate edit_freq to get equal column number to  
  he_rse <- calc_edit_frequency(he_rse)

  n_assays <- length(assays(nh_rse))
  assay_nms <- assayNames(nh_rse)
  new_assays <- lapply(seq_len(n_assays), function(i) {
    do.call(rbind, list(assays(nh_rse)[[i]], 
                        assays(rp_he_rse)[[i]], 
                        assays(he_rse)[[i]]))
  })
  
  names(new_assays) <- assay_nms
  
  new_rowRanges <- c(rowRanges(nh_rse), rowRanges(rp_he_rse), rowRanges(he_rse))
  new_colData <- colData(nh_rse)
  rse <- SummarizedExperiment(assays = new_assays,
                              colData = new_colData,
                              rowRanges = new_rowRanges)
  he_rse <- calc_edit_frequency(he_rse)
  saveRDS(rse, "rse-final.rds")
} else {
  rse <- readRDS("rse-final.rds")
}

he_rse

rse
```

```{r}
alt <- assay(rse, "nG")
ref <- assay(rse, "nA")
edit_idx <- colSums(alt) / (colSums(alt) + colSums(ref))
edit_idx <- data.frame(sample_id = factor(names(edit_idx), 
                                          levels = levels(sample_info$sample_id)),
                       edit_idx = edit_idx)
edit_idx$condition <-  factor(sub("_[0-9]$", "", as.character(edit_idx$sample_id)), levels = sample_order)

edit_idx
```

```{r}
pa <- ggplot(edit_idx, aes(condition, edit_idx)) + 
  geom_jitter(width = 0.2) + 
  stat_summary(
    fun.y = "mean",
    col = "red",
    size = 0.2,
    shape = 3
  ) + 
  labs(
    x = NULL,
    y = "Editing index (including HE sites)") +
  theme(axis.text.x =element_text(angle = 90, hjust = 1, vjust = 0.5))
pa

p <- ggplot(edit_idx[edit_idx$condition %in% subset_to_plot,  ], aes(condition, edit_idx)) + 
  geom_jitter(width = 0.1) + 
  stat_summary(
    fun.y = "mean",
    col = "red",
    size = 0.2,
    shape = 3
  ) + 
  labs(
    x = NULL,
    y = "Editing index (including HE sites)") +
  theme(axis.text.x =element_text(angle = 90, hjust = 1, vjust = 0.5))

save_plot("plots/chk_AEI_final.pdf", p, base_asp = 1, base_height = 4)

p
```

```{r}
rse <- readRDS("rse-final.rds")

rse$condition <- sub("_[0-9]$", "", as.character(rse$sample))
rse$condition <- factor(rse$condition, levels = sample_order)
rse$sample <- factor(rse$sample)

if(!file.exists("differential_editing_results.rds")) {
  conds <- list(
    c("ADAR1p110-replica", "ADAR1p150-replica"),
    c("ADAR1p150-replica", "RFP_editase_neg_ctrl"),
    c("ADAR1p110-replica", "RFP_editase_neg_ctrl")
  )

  de_res_all <- lapply(conds, function(x) {
    message(x)
    filt_rse <- rse[, rse$condition %in% x]
    filt_rse$sample <- droplevels(filt_rse$sample)
    filt_rse$condition <- droplevels(filt_rse$condition)
    has_depth <- rowSums(assay(filt_rse, "depth") >= 3) == ncol(filt_rse)
    has_edits <- rowSums(assay(filt_rse, "edit_freq") >= 0.05) >= 3
    filt_rse <- filt_rse[has_depth & has_edits, ]
    deobj <- make_de_object(filt_rse)
    de_results <- find_de_sites(deobj,
                                test = "DESeq2",
                                sample_col = "sample",
                                condition_col = "condition",
                                condition_control = x[2],
                                condition_treatment = x[1]
    )
  })

    de_res_all <- lapply(de_res_all, function(x) {
    xx <- x$results_full
    xx <- xx[which(xx$padj < 0.05), ]
    xx <- xx[order(xx$pvalue), ]
    x$sig_results <- as.data.frame(xx)
    x
  })
  names(de_res_all) <- sapply(conds, paste, collapse = "_vs_")
  saveRDS(de_res_all, "differential_editing_results.rds")
} else {
    de_res_all <- readRDS("differential_editing_results.rds")
}

n_sig_sites <- sapply(de_res_all, function(x) x$sig_results |> nrow())
sig_sites <- data.frame(comparision = names(n_sig_sites), significant_sites = n_sig_sites,
           row.names = NULL)

ggplot(sig_sites, aes(comparision, significant_sites)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  theme_minimal() +
  labs(x = "comparision", y = "# of significant sites")

```


```{r, eval = FALSE}
# use limma-voom instead for performance
library(edgeR)
has_depth <- rowSums(assay(rse, "depth") >= 10) >= 3
has_depth <- rowSums(assay(rse, "depth") >= 5) == ncol(rse)

has_edits <- rowSums(assay(rse, "edit_freq") >= 0.05) >= 3

filt_rse <- rse[has_depth & has_edits, ]
deobj <- make_de_object(filt_rse)

#design <- model.matrix(~ 0 + sample + condition:count)
# Provide the sample metadata 
design <- model.matrix(~ 0 + sample + condition:count, data = colData(deobj))

design <- design[, !grepl("countref", colnames(design))]
mod_mat <- design

dge <- edgeR::DGEList(assay(deobj, "counts"),
                      lib.size = rep(1, ncol(deobj))
)
#######Variance checkplot
y <- voom(dge, mod_mat, plot = T)
fit <- lmFit(y, mod_mat)
fit
lv_res <- lapply(conds, function(to_test) {
  condition_treatment <- to_test[1]
  condition_control <- to_test[2]
  alt_treatment <- colMeans(design[deobj$condition == condition_treatment &
                                     deobj$count == "alt", ])
  ref_treatment <- colMeans(design[deobj$condition == condition_treatment &
                                     deobj$count == "ref", ])

  alt_control <- colMeans(design[deobj$condition == condition_control &
                                   deobj$count == "alt", ])
  ref_control <- colMeans(design[deobj$condition == condition_control &
                                   deobj$count == "ref", ])

  contrast_to_test <- (alt_treatment - ref_treatment) - (alt_control - ref_control)

  tmp <- contrasts.fit(fit, contrast_to_test)
  tmp <- eBayes(tmp)
  tt <- topTable(tmp, sort.by = "P", n = Inf)
  tt
})

lv_res <- lapply(lv_res, function(x) {
  xx <- list()
  xx$results_full <- x

  x <- x[which(x$adj.P.Val < 0.05), ]
  x <- x[order(x$P.Value), ]
  xx$sig_results <- as.data.frame(x)
  xx
})

names(lv_res) <- sapply(conds, paste, collapse = "_vs_")

sapply(lv_res, function(x) nrow(x$sig_results))

```

## Visualize p150 specific editing sites

```{r}
plot_editing_counts <- function(rse,
                                site = "site_X",
                                samples = colnames(rse),
                                group = "condition") {
  ef <- 100 * assay(rse, "edit_freq")[site, ]
  dp <- assay(rse, "depth")[ site, ]
  site_id <- as.character(rowRanges(rse)[site, ])
  plt_dat <- data.frame(editing_frequency = ef,
                        sample = colnames(rse),
                        depth = dp,
                        condition = rse[[group]])
    
  ggplot(plt_dat, aes(condition, editing_frequency)) +
    geom_jitter() +
    labs(y = "Percent editing",
         x = NULL,
         subtitle = site_id) +
    theme(axis.text.x = element_text(angle = 90,
                                     hjust = 1, vjust = 0.5))
}

plot_editing_heatmap <- function(x, sites, scale = FALSE, ...) {
  ef <- assays(x)$edit_freq[sites, ]
  if(scale) {
    ef <- t(scale(t(ef)))
  }
  colnames(ef) <- colData(x)$condition
  Heatmap(ef,...)
}



plot_editing_heatmap <- function(x, sites, scale = FALSE, ...) {
  ef <- assays(x)$edit_freq[sites, ]
  if(scale) {
    ef <- t(scale(t(ef)))
  }
  colnames(ef) <- colData(x)$condition
  Heatmap(ef,...)
}

plot_editing_heatmap <- function(x, sites, scale = FALSE, ...) {
  ef <- assays(x)$edit_freq[sites, ]
  ef[is.infinite(ef)] <- NA
  ef <- ef[rowSums(is.na(ef)) < ncol(ef), , drop = FALSE]
  ef[is.na(ef)] <- 0
  keep <- apply(ef, 1, function(r) var(r) > 0)
  ef <- ef[keep, , drop = FALSE]
  if (scale) {
    ef <- t(scale(t(ef)))
    ef[is.na(ef)] <- 0
  }
  colnames(ef) <- colData(x)$condition
  Heatmap(ef,...)
}
```

```{r, fig.width = 12, fig.height= 12}
p150_sites <- de_res_all$`ADAR1p110-replica_vs_ADAR1p150-replica`$sig_results |>
  rownames() 

cols_to_keep <- rse$condition %in% subset_to_plot 
```

```{r, fig.width = 12, fig.height= 12}
p <- plot_editing_heatmap(rse[, cols_to_keep],
                     p150_sites,
                     show_row_names = FALSE, name = "Editing Frequency",
                     column_title = paste0("p150 dependent editing sites, (n = ", 
                                           length(p150_sites), 
                                           ")"))

pp <- plot_editing_heatmap(rse[, cols_to_keep],
                     p150_sites, 
                     show_row_names = FALSE,
                     scale = TRUE, 
                     name = "Z-score of\n Editing Frequency",
                     column_title = paste0("p150 dependent editing sites, (n = ", 
                                           length(p150_sites), 
                                           ")"))
                                           
p
pp 

pdf("plots/p150_sites.pdf", width = 8, height = 10)
print(p)
dev.off()
         
pdf("plots/p150_sites_zscore.pdf", width = 8, height = 10)
print(pp)
dev.off()         
                                           
```

```{r, eval=FALSE, fig.cap="Editing frequency (# of edited reads / total reads) of p110  sites summed across each sample."}
alt <- assay(rse, "nG")[p150_sites, ]
ref <- assay(rse, "nA")[p150_sites, ]
edit_idx <- colSums(alt) / (colSums(alt) + colSums(ref))
edit_idx <- data.frame(sample_id = factor(names(edit_idx), 
                                          levels = levels(sample_info$sample_id)),
                       edit_idx = edit_idx)
edit_idx$condition <-  factor(sub("_[0-9]$", "", as.character(edit_idx$sample_id)), levels = sample_order)


pa <- ggplot(edit_idx, aes(condition, edit_idx)) + 
  geom_jitter(width = 0.2) + 
  stat_summary(
    fun.y = "mean",
    col = "red",
    size = 0.2,
    shape = 3
  ) + 
  labs(
    x = NULL,
    y = "Editing index (p150 sites)") +
  theme(axis.text.x =element_text(angle = 90, hjust = 1, vjust = 0.5))
pa

save_plot("plots/p150_sites_idx.pdf", pa, base_asp = 1, base_height = 4)
```
Next i'll plot the top 12 editing sites differentially edited 

Notice how the pattern in these individual sites mirrors the global pattern seen in the heatmaps and overall editing indexes. Namely, the delta z alpha and AGS mutations very subtly reduce editing levels compared to the p150_plus_Dox controls. 

```{r, fig.width = 12, fig.height = 18}
plts <- lapply(p150_sites[1:12], plot_editing_counts, rse = rse)
plot_grid(plotlist = plts, nrow = 4, ncol = 3)

grid_plot <- plot_grid(plotlist = plts, nrow = 4, ncol = 3)

ggsave("plots/p150_sites_pgrid.pdf", plot = grid_plot, width = 12, height = 18, units = "in")

```

```{r}
library(openxlsx)
ro <- lapply(de_res_all, "[[", "sig_results") |> sapply(nrow) |> order(decreasing = TRUE)
out <- lapply(de_res_all[ro], function(x) {
  xx <- x$sig_results
  edit_counts <- counts(x$de_obj)[rownames(xx), ]
  res <- cbind(site = rownames(xx), xx, edit_counts)
  res
})
wb <- createWorkbook()
for(i in seq_along(out)) {
  addWorksheet(wb, substr(names(out)[i], 1, 31))
  writeData(wb, i, paste("Comparision is:", names(out)[i]), 1, 1)
  writeDataTable(wb, i, out[[i]], startCol = 1, startRow = 2)
}
saveWorkbook(wb, "differential_editing_results.xlsx", overwrite = TRUE)

```

## Session info

<details style="margin-bottom:10px;">
<summary>
Session info
</summary>

```{r}
sessionInfo()
```


</details>



